\pkg{depmixS4}: An \proglang{R} Package for Hidden Markov Models

### 2010-10-18 papers/jss/depmixS4.Rnw
Minor typo's corrected.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@440 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-

\documentclass[article]{jss}
\usepackage{amsmath}

%\usepackage[]{amsmath, amsfonts, amstext, amsthm} 
%\usepackage{amssymb}
%\usepackage[]{graphics} 
%\usepackage{epsfig}
%\usepackage{epstopdf}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Ingmar Visser\\University of Amsterdam \And 
        Maarten Speekenbrink\\University College London}

\title{\pkg{depmixS4} : An \proglang{R}-package for Hidden Markov Models}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated
\Plaintitle{depmixS4: An R-package for Hidden Markov Models} %% without formatting

\Shorttitle{depmixS4: Hidden Markov Models} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{

	\pkg{depmixS4} implements a general framework for defining and
	estimating dependent mixture models in the \proglang{R}
	programming language \citep{R2010}.  This includes standard Markov
	models, latent/hidden Markov models, and latent class and finite
	mixture distribution models.  The models can be fitted on mixed
	multivariate data with distributions from the \code{glm} family,
	the (logistic) multinomial, or the multivariate normal
	distribution.  Other distributions can be added easily, and an
	example is provided with the {\em exgaus} distribution.
	Parameters are estimated by the EM algorithm or, when (linear)
	constraints are imposed on the parameters, by direct numerical
	optimization with the \pkg{Rsolnp} or \pkg{Rdonlp2} routines.

}

@@

\Plainkeywords{hidden Markov model, dependent mixture model, mixture 
model, constraints} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{13}
%% \Issue{9}
%% \Month{September}
%% \Year{2004}
%% \Submitdate{2004-09-29}
%% \Acceptdate{2004-09-29}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Ingmar Visser\\
Department of Psychology\\
University of Amsterdam\\
Roetersstraat 15\\
1018 WB, Amsterdam\\
The Netherlands\\
E-mail: \email{i.visser@uva.nl} \\
URL: \url{http://www.ingmar.org/}
}

%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/1/31336-5053
%% Fax: +43/1/31336-734

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\batchmode

\SweaveOpts{echo=TRUE}
\usepackage{a4wide}

%% \usepackage{Sweave}

@@

%set width of figures produced by Sweave
\setkeys{Gin}{width=0.8\textwidth}

\maketitle

<<echo=FALSE,results=hide>>=	
options(prompt="R> ", continue="+	",width=70, useFancyQuotes=FALSE,
digits=4)
library(depmixS4)
@

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\section{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes; see \citet{Wickens1982},
for an overview, and e.g. \citet{Schmittmann2006}, for a recent
application.  In economics, latent Markov models are so-called regime
switching models (see e.g., \citealp{Kim1994} and
\citealp{Ghysels1994}).  Further applications include speech
recognition \citep{Rabiner1989}, EEG analysis \citep{Rainer2000}, and
genetics \citep{Krogh1998}.  In these latter areas of application,
latent Markov models are usually referred to as hidden Markov models.
See for example \citet{Fruhwirth2006} for an overview of hidden Markov
models with extensions.  Further examples of applications can be found
in e.g.\ \citet[][chapter~1]{Cappe2005}.  A more gentle introduction
into hidden Markov models with applications is the book by
\citet{Zucchini2009}.

The \pkg{depmixS4} package was motivated by the fact that while Markov
models are used commonly in the social sciences, no comprehensive
package was available for fitting such models.  Existing software for
estimating Markovian models include Panmark \citep{Pol1996}, and for
latent class models Latent Gold \citep{Vermunt2003}.  These programs
lack a number of important features, besides not being freely
available.  There are currently some packages in \proglang{R} that
handle hidden Markov models but they lack a number of features that we
needed in our research.  In particular, \pkg{depmixS4} was designed to
meet the following goals:

\begin{enumerate}

	\item to be able to estimate parameters subject to general
	linear (in)equality constraints

	\item to be able to fit transition models with covariates, i.e.,
	to have time-dependent transition matrices

	\item to be able to include covariates in the prior or initial
	state probabilities

	\item to be easily extensible, in particular, to allow users to
	easily add new uni- or multivariate response distributions and
	new transition models, e.g., continuous time observation models

@@

Although \pkg{depmixS4} was designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case when
$T=1$.  In this case, there are no time dependencies between observed
data and the model reduces to a finite mixture or latent class model.
While there are specialized packages to deal with mixture data, as far
as we know these don't allow the inclusion of covariates on the prior
probabilities of class membership.  The possibility to estimate the
effects of covariates on prior and transition probabilities is a
distinguishing feature of \pkg{depmixS4}.  In the next section, we
provide an outline of the model and likelihood equations.

\section{The dependent mixture model}

@@

\begin{figure}[!Htbp]
\begin{center}
<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=	
data(speed)
plot(as.ts(speed[1:168,]),main="Speed-accuracy trade-off")
@
	\caption{Response times (rt), accuracy (corr) and pay-off 
	values (Pacc) for the first series of responses in dataset 
	\code{speed}.}
	\label{fig:speed}
\end{center}
\end{figure}

The latent Markov model is usually associated with data of this type,
in particular for multinomially distributed responses.  However,
commonly employed estimation procedures \citep[e.g.,][]{Pol1996}, are
not suitable for long time series due to underflow problems.  In
contrast, the hidden Markov model is typically only used for `long'
univariate time series \citep[][chapter~1]{Cappe2005}.  We use the
term "dependent mixture model" because one of the authors (Ingmar
Visser) thought it was time for a new name to relate these
models\footnote{Only later did I find out that \citet{Leroux1992}
already coined the term dependent mixture models in an application
with hidden Markov mixtures of Poisson count data.}.

@@

	\item $a_{ij}(\vc{z}_t) = \Prob(S_{t+1}=j|S_{t}=i,\vc{z}_t)$,
	provides the probability of a transition from state $i$ to state
	$j$ with covariate $\vc{z}_t$,

@@

To obtain maximum likelihood estimates of the model parameters, we
need the marginal likelihood of the observations.  For hidden Markov
models, this marginal (log-)likelihood is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \cite{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters and covariates is dropped here):
\begin{equation}
	L_{T} = \Prob(\vc{O}_{1:T}) = \prod_{t=1}^{T} 
	\Prob(\vc{O}_{t}|\vc{O}_{1:(t-1)}), 
	\label{condLike}
\end{equation}
where $\Prob(\vc{O}_{1}|\vc{O}_{0}):=\Prob(\vc{O}_{1})$. Note that for a 
simple, i.e.\ observed, Markov chain these probabilities reduce to 
$\Prob(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=\Prob(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[\Prob(\vc{O}_{t}|\vc{O}_{1:(t-1)}].
	\label{eq:condLogl}
\end{equation}

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM
algorithm, parameters are estimated by iteratively maximizing the
expected joint log-likelihood of the parameters given the observations and
states.  Let $\greekv{\theta} = (\greekv{\theta}_1,
\greekv{\theta}_2,\greekv{\theta}_3)$ be the general parameter vector
consisting of three subvectors with parameters for the prior model,
transition model, and response models respectively.  The joint
log-likelihood can be written as:
\begin{equation}
\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
Expectation step, we replace these with their expected values given a set of 
(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $\vc{O}_{1:T}$. 
The expected log-likelihood:
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta})),
\end{equation}
can be written as:
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\log \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
\end{multline}
where the expected values $\xi_t(j,k) = P(S_t = k, S_{t-1} =
j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) =
P(S_t = j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ can be
computed effectively by the forward-backward algorithm \citep[see
e.g.,][]{Rabiner1989}.  The Maximization step consists of the
maximization of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
side of (\ref{eq:Q}) consists of three separate parts, we can maximize
separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and
$\greekv{\theta}_3$.  In common models, maximization for
$\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the
\code{nnet.default} routine in the \pkg{nnet} package
\citep{Venables2002}, and maximization for $\greekv{\theta}_3$ by the
standard \code{glm} routine.  Note that for the latter maximization,
the expected values $\gamma_t(j)$ are used as prior weights of the
observations $O^k_t$.

@@

	\item model specification with function \code{depmix} (or with
	\code{mix} for latent class and finite mixture models, see example
	below on adding covariates to prior probabilities)

	\item  model fitting with function \code{fit}
\end{enumerate}
We have separated the stages of model specification and model fitting
because fitting large models can be fairly time-consuming and it is
hence useful to be able to check the model specification before
actually fitting the model.

@@

Throughout this article a data set called \code{speed} is used.  As
already indicated in the Introduction, it consists of three time
series with three variables: response time, accuracy, and a covariate,
Pacc, which defines the relative pay-off for speeded versus accurate
responding.  Before describing some of the models that are fitted to
these data, we provide a brief sketch of the reasons for gathering
these data in the first place.

@@

One potential problem with the diffusion model is that it predicts a
continuous trade-off between speed and accuracy of responding, i.e.,
when participants are pressed to respond faster and faster, the
diffusion model predicts that this would lead to a gradual decrease in
accuracy.  The \code{speed} data set that we analyze below was
gathered to test this hypothesis versus the alternative hypothesis
stating that there is a sudden transition from slow and accurate
responding to fast responding at chance level.  At each trial of the
experiment, the participant is shown the current setting of the
relative reward for speed versus accuracy.  The bottom panel of
figure~\ref{fig:speed} shows the values of this variable.  The
experiment was designed to investigate what would happen when this
reward variable changes from reward for accuracy only to reward for
speed only.  The \code{speed} data that we analyse here are from
participant A in Experiment 1 in \citet{Dutilh2010}, who provide a
complete description of the experiment and the relevant theoretical
background.

@@

A dependent mixture model is defined by the number of states and the
initial state, state transition, and response distribution functions.
A dependent mixture model can be created with the
\code{depmix}-function as follows:
<<>>=
library(depmixS4)
data(speed)
set.seed(1)
mod <- depmix(response=rt~1, data=speed, nstates=2, trstart=runif(4))
@

@@

The call to \code{depmix} specifies the model with a number of
arguments.  The \code{response} argument is used to specify the
response variable, possibly with covariates, in the familiar format
using formulae such as in \code{lm} or \code{glm} models.  The second
argument, \code{data=speed}, provides the \code{data.frame} in which
the variables from \code{response} are interpreted.  Third, the number
of states of the model is given by \code{nstates=2}.

\paragraph{Starting values} Note also that start values for the
transition parameters are provided in this call by the \code{trstart}
argument.  Starting values for parameters can be provided using three
arguments: \code{instart} for the parameters relating to the prior
probabilities, \code{trstart}, relating the transition models, and
\code{respstart} for the parameters of the response models.  Note that
the starting values for the initial and transition models as well as
multinomial logit response models are interpreted as {\em
probabilities}, and internally converted to multinomial logit
parameters (if a logit link function is used).  Start values can also
be generated randomly within the EM algorithm by generating random
uniform values for the values of $\gamma_{t}$ in (\ref{eq:Q}) at
iteration 0.  Automatic generation of starting values is not available
for constrained models (see below).

\paragraph{Fitting the model and printing results} The \code{depmix}
function returns an object of class \code{depmix} which contains the
model specification, and not a fitted model.  Hence, the model needs
to be fitted by calling \code{fit}:
<<>>=
fm <- fit(mod)
@

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The class has \code{print} and
\code{summary} methods to see the results.  The \code{print} method
provides information on convergence, the log-likelihood and
the AIC and BIC values:
<<>>=
fm 
@

These statistics can also be extracted using \code{logLik}, \code{AIC}
and \code{BIC}, respectively.  By comparison, a 1-state model for
these data, i.e. assuming there is no mixture, has a log-likelihood of
$-305.33$, and 614.66, and 622.83 for the AIC and BIC respectively.
Hence, the 2-state model fits the data much better than the 1-state
model.  Note that the 1-state model can be specified using \code{mod
<- depmix(rt~1, data=speed, nstates=1)}, although this model is
trivial as it will simply return the mean and standard deviation of
the \code{rt} variable.

@@

By default, the transition probabilities and the initial state
probabilities are parameterized using a multinomial model with an
identity link function.  Using a multinomial logistic model allows one
to include covariates on the initial state and transition
probabilities.  In this case, each row of the transition matrix is
parameterized by a baseline category logistic multinomial, meaning
that the parameter for the base category is fixed at zero
\citep[see][p.\ 267 ff., for multinomial logistic models and various
parameterizations]{Agresti2002}.  The default baseline category is the
first state.  Hence, for example, for a 3-state model, the initial
state probability model would have three parameters of which the first
is fixed at zero and the other two are freely estimated.
\citet{Chung2007} discuss a related latent transition model for
repeated measurement data ($T=2$) using logistic regression on the
transition parameters; they rely on Bayesian methods of estimation.
Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
<<>>=
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(),
	   transition=~scale(Pacc), instart=runif(2)) 
fm <-fit(mod,verbose=FALSE) 
@

Note the use of \code{verbose=FALSE} to suppress printing of
information on the iterations of the fitting process.  Applying
\code{summary} to the \code{fit}ted object gives (only transition
models printed here by using argument \code{which}): 
<<>>= 
summary(fm, which="transition") 
@

@@

Multivariate data can be modelled by providing a list of formulae as
well as a list of family objects for the distributions of the various
responses.  In above examples we have only used the response times
which were modelled as a Gaussian distribution.  The accuracy
variable in the \code{speed} data can be modelled with a multinomial
by specifying the following:
<<>>=
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
    family=list(gaussian(), multinomial("identity")),
		transition=~scale(Pacc),instart=runif(2))
fm <- fit(mod,verbose=FALSE)
@

This provides the following fitted model parameters (only the 
response parameters are given here): 
<<>>=
summary(fm,which="response")
@

@@

\begin{figure}[htbp]
\begin{center}
	\includegraphics{graphs/baldist.pdf}
	\caption{Balance scale item; this is a distance item (see the text 
	for details).}
	\label{fig:balance}  
\end{center}
\end{figure}

>>>>

\documentclass[article,nojss]{jss}
\usepackage{amsmath,thumbpdf}
\graphicspath{{Figures/}}

%% \usepackage{Sweave}
\SweaveOpts{engine = R, echo = TRUE, keep.source = TRUE, eps = FALSE}

%\VignetteIndexEntry{depmixS4: An R Package for Hidden Markov Models}
%\VignetteDepends{depmixS4,gamlss,gamlss.data}
%\VignetteKeywords{hidden Markov model, dependent mixture model, mixture model, constraints}
%\VignettePackage{depmixS4}

\author{Ingmar Visser\\University of Amsterdam \And 
        Maarten Speekenbrink\\University College London}
\Plainauthor{Ingmar Visser, Maarten Speekenbrink}

\title{\pkg{depmixS4}: An \proglang{R} Package for Hidden Markov Models}
\Plaintitle{depmixS4: An R Package for Hidden Markov Models}

\Abstract{	
  This introduction to the \proglang{R} package \pkg{depmixS4} is a (slightly)
  modified version of \cite{Visser+Speekenbrink:2010}, published in the
  \emph{Journal of Statistical Software}.

  \pkg{depmixS4} implements a general framework for defining and
  estimating dependent mixture models in the \proglang{R}
  programming language.  This includes standard Markov
  models, latent/hidden Markov models, and latent class and finite
  mixture distribution models.  The models can be fitted on mixed
  multivariate data with distributions from the \code{glm} family,
  the (logistic) multinomial, or the multivariate normal
  distribution.  Other distributions can be added easily, and an
  example is provided with the {\em exgaus} distribution.
  Parameters are estimated by the expectation-maximization (EM) algorithm or, when (linear)
  constraints are imposed on the parameters, by direct numerical
  optimization with the \pkg{Rsolnp} or \pkg{Rdonlp2} routines.   
}

@@

\Volume{36}
\Issue{7}
\Month{August}
\Year{2010}
\Submitdate{2009-08-19}
\Acceptdate{2010-06-21}

\Address{
  Ingmar Visser\\
  Department of Psychology\\
  University of Amsterdam\\
  Roetersstraat 15\\
  1018 WB, Amsterdam, The Netherlands\\
  E-mail: \email{i.visser@uva.nl} \\
  URL: \url{http://www.ingmar.org/}
}

@@

<<echo=FALSE,results=hide>>=	
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE, digits = 4)
library("depmixS4")
@

\section{Introduction}

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes; see \citet{Wickens1982},
for an overview, and e.g., \citet{Schmittmann2006}, for a recent
application.  In economics, latent Markov models are so-called regime
switching models (see e.g., \citealp{Kim1994} and
\citealp{Ghysels1994}).  Further applications include speech
recognition \citep{Rabiner1989}, EEG analysis \citep{Rainer2000}, and
genetics \citep{Krogh1998}.  In these latter areas of application,
latent Markov models are usually referred to as hidden Markov models.
See for example \citet{Fruhwirth2006} for an overview of hidden Markov
models with extensions.  Further examples of applications can be found
in e.g., \citet[][Chapter~1]{Cappe2005}.  A more gentle introduction
into hidden Markov models with applications is the book by
\citet{Zucchini2009}.

The \pkg{depmixS4} package was motivated by the fact that while Markov
models are used commonly in the social sciences, no comprehensive
package was available for fitting such models.  Existing software for
estimating Markovian models include \pkg{Panmark} \citep{Pol1996}, and for
latent class models \pkg{Latent Gold} \citep{Vermunt2003}.  These programs
lack a number of important features, besides not being freely
available.  There are currently some packages in \proglang{R} that
handle hidden Markov models but they lack a number of features that we
needed in our research.  In particular, \pkg{depmixS4} was designed to
meet the following goals:

\begin{enumerate}	
	\item to be able to estimate parameters subject to general
	linear (in)equality constraints;

	\item to be able to fit transition models with covariates, i.e.,
	to have time-dependent transition matrices;

	\item to be able to include covariates in the prior or initial
	state probabilities;

	\item to be easily extensible, in particular, to allow users to
	easily add new uni- or multivariate response distributions and
	new transition models, e.g., continuous time observation models.

@@

Although \pkg{depmixS4} was designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case when
$T=1$.  In this case, there are no time dependencies between observed
data and the model reduces to a finite mixture or latent class model.
While there are specialized packages to deal with mixture data, as far
as we know these don't allow the inclusion of covariates on the prior
probabilities of class membership.  The possibility to estimate the
effects of covariates on prior and transition probabilities is a
distinguishing feature of \pkg{depmixS4}.
In Section~\ref{sec:2}, we provide an outline of the model and likelihood equations.

The \pkg{depmixS4} package is implemented using the \proglang{R} system
for statistical computing \citep{R2010} and is available from the
Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/package=depmixS4}.

\section{The dependent mixture model} \label{sec:2}

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[t!]
\begin{center}
<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=	
data("speed")
plot(as.ts(speed[1:168,]), main = "Speed-accuracy trade-off")
@
	\caption{Response times (rt), accuracy (corr) and pay-off 
	values (Pacc) for the first series of responses in dataset 
	\code{speed}.}
	\label{fig:speed}
\end{center}
\end{figure}

The latent Markov model is usually associated with data of this type,
in particular for multinomially distributed responses.  However,
commonly employed estimation procedures \citep[e.g.,][]{Pol1996}, are
not suitable for long time series due to underflow problems.  In
contrast, the hidden Markov model is typically only used for `long'
univariate time series \citep[][Chapter~1]{Cappe2005}.  We use the
term ``dependent mixture model" because one of the authors (Ingmar
Visser) thought it was time for a new name to relate these
models\footnote{Only later he found out that \citet{Leroux1992}
already coined the term dependent mixture models in an application
with hidden Markov mixtures of Poisson count data.}.

@@

	\item $a_{ij}(\vc{z}_t) = \Prob(S_{t+1}=j|S_{t}=i,\vc{z}_t)$,
	provides the probability of a transition from state $i$ to state
	$j$ with covariate $\vc{z}_t$.

@@

To obtain maximum likelihood estimates of the model parameters, we
need the marginal likelihood of the observations.  For hidden Markov
models, this marginal (log-)likelihood is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \cite{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters and covariates is dropped here):
\begin{equation}
	L_{T} = \Prob(\vc{O}_{1:T}) = \prod_{t=1}^{T} 
	\Prob(\vc{O}_{t}|\vc{O}_{1:(t-1)}), 
	\label{condLike}
\end{equation}
where $\Prob(\vc{O}_{1}|\vc{O}_{0}):=\Prob(\vc{O}_{1})$. Note that for a 
simple, i.e., observed, Markov chain these probabilities reduce to 
$\Prob(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=\Prob(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[\Prob(\vc{O}_{t}|\vc{O}_{1:(t-1)}].
	\label{eq:condLogl}
\end{equation}

@@

Parameters are estimated in \pkg{depmixS4} using the expectation-maximization (EM) algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM
algorithm, parameters are estimated by iteratively maximizing the
expected joint log-likelihood of the parameters given the observations and
states.  Let $\greekv{\theta} = (\greekv{\theta}_1,
\greekv{\theta}_2,\greekv{\theta}_3)$ be the general parameter vector
consisting of three subvectors with parameters for the prior model,
transition model, and response models respectively.  The joint
log-likelihood can be written as:
\begin{multline}
\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) \\
+ \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
\end{multline}
This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
Expectation step, we replace these with their expected values given a set of 
(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $\vc{O}_{1:T}$. 
The expected log-likelihood:
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta})),
\end{equation}
can be written as:
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\log \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
\end{multline}
where the expected values $\xi_t(j,k) = P(S_t = k, S_{t-1} =
j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) =
P(S_t = j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ can be
computed effectively by the forward-backward algorithm \citep[see
e.g.,][]{Rabiner1989}.  The Maximization step consists of the
maximization of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
side of (\ref{eq:Q}) consists of three separate parts, we can maximize
separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and
$\greekv{\theta}_3$.  In common models, maximization for
$\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the
\code{nnet.default} routine in the \pkg{nnet} package
\citep{Venables2002}, and maximization for $\greekv{\theta}_3$ by the
standard \code{glm} routine.  Note that for the latter maximization,
the expected values $\gamma_t(j)$ are used as prior weights of the
observations $O^k_t$.

@@

	\item model specification with function \code{depmix} (or with
	\code{mix} for latent class and finite mixture models, see example
	below on adding covariates to prior probabilities);

	\item  model fitting with function \code{fit}.
\end{enumerate}
We have separated the stages of model specification and model fitting
because fitting large models can be fairly time-consuming and it is
hence useful to be able to check the model specification before
actually fitting the model.

@@

Throughout this article a data set called \code{speed} is used.  As
already indicated in the introduction, it consists of three time
series with three variables: response time \code{rt}, accuracy
\code{corr}, and a covariate, \code{Pacc}, which defines the relative
pay-off for speeded versus accurate responding.  Before describing
some of the models that are fitted to these data, we provide a brief
sketch of the reasons for gathering these data in the first place.

@@

One potential problem with the diffusion model is that it predicts a
continuous trade-off between speed and accuracy of responding, i.e.,
when participants are pressed to respond faster and faster, the
diffusion model predicts that this would lead to a gradual decrease in
accuracy.  The \code{speed} data set that we analyze below was
gathered to test this hypothesis versus the alternative hypothesis
stating that there is a sudden transition from slow and accurate
responding to fast responding at chance level.  At each trial of the
experiment, the participant is shown the current setting of the
relative reward for speed versus accuracy.  The bottom panel of
Figure~\ref{fig:speed} shows the values of this variable.  The
experiment was designed to investigate what would happen when this
reward variable changes from reward for accuracy only to reward for
speed only.  The \code{speed} data that we analyse here are from
participant A in Experiment 1 in \citet{Dutilh2010}, who provide a
complete description of the experiment and the relevant theoretical
background.

@@

A dependent mixture model is defined by the number of states and the
initial state, state transition, and response distribution functions.
A dependent mixture model can be created with the
\code{depmix} function as follows:
<<>>=
library("depmixS4")
data("speed")
set.seed(1)
mod <- depmix(response = rt ~ 1, data = speed, nstates = 2,
  trstart = runif(4))
@

@@

The call to \code{depmix} specifies the model with a number of
arguments.  The \code{response} argument is used to specify the
response variable, possibly with covariates, in the familiar format
using formulae such as in \code{lm} or \code{glm} models.  The second
argument, \code{data = speed}, provides the \code{data.frame} in which
the variables from \code{response} are interpreted.  Third, the number
of states of the model is given by \code{nstates = 2}.

\paragraph{Starting values.} Note also that start values for the
transition parameters are provided in this call by the \code{trstart}
argument.  Starting values for parameters can be provided using three
arguments: \code{instart} for the parameters relating to the prior
probabilities, \code{trstart}, relating the transition models, and
\code{respstart} for the parameters of the response models.  Note that
the starting values for the initial and transition models as well as
multinomial logit response models are interpreted as {\em
probabilities}, and internally converted to multinomial logit
parameters (if a logit link function is used).  Start values can also
be generated randomly within the EM algorithm by generating random
uniform values for the values of $\gamma_{t}$ in (\ref{eq:Q}) at
iteration 0.  Automatic generation of starting values is not available
for constrained models (see below).

\paragraph{Fitting the model and printing results.} The \code{depmix}
function returns an object of class `\code{depmix}' which contains the
model specification, and not a fitted model.  Hence, the model needs
to be fitted by calling \code{fit}:
<<>>=
fm <- fit(mod)
@

The \code{fit} function returns an object of class
`\code{depmix.fitted}' which extends the `\code{depmix}' class, adding
convergence information (and information about constraints if these
were applied, see below).  The class has \code{print} and
\code{summary} methods to see the results.  The \code{print} method
provides information on convergence, the log-likelihood and
the AIC and BIC values:
<<>>=
fm 
@

These statistics can also be extracted using \code{logLik}, \code{AIC}
and \code{BIC}, respectively.  By comparison, a 1-state model for
these data, i.e., assuming there is no mixture, has a log-likelihood of
$-305.33$, and 614.66, and 622.83 for the AIC and BIC respectively.
Hence, the 2-state model fits the data much better than the 1-state
model.  Note that the 1-state model can be specified using
\code{mod <- depmix(rt ~ 1, data = speed, nstates = 1)}, although this model is
trivial as it will simply return the mean and standard deviation of
the \code{rt} variable.

@@

By default, the transition probabilities and the initial state
probabilities are parameterized using a multinomial model with an
identity link function.  Using a multinomial logistic model allows one
to include covariates on the initial state and transition
probabilities.  In this case, each row of the transition matrix is
parameterized by a baseline category logistic multinomial, meaning
that the parameter for the base category is fixed at zero
\citep[see][p.~267~ff., for multinomial logistic models and various
parameterizations]{Agresti2002}.  The default baseline category is the
first state.  Hence, for example, for a 3-state model, the initial
state probability model would have three parameters of which the first
is fixed at zero and the other two are freely estimated.
\citet{Chung2007} discuss a related latent transition model for
repeated measurement data ($T=2$) using logistic regression on the
transition parameters; they rely on Bayesian methods of estimation.
Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
<<>>=
set.seed(1)
mod <- depmix(rt ~ 1, data = speed, nstates = 2, family = gaussian(),
  transition = ~ scale(Pacc), instart = runif(2)) 
fm <- fit(mod, verbose = FALSE) 
@

Note the use of \code{verbose = FALSE} to suppress printing of
information on the iterations of the fitting process.  Applying
\code{summary} to the \code{fit}ted object gives (only transition
models printed here by using argument \code{which}): 
<<>>= 
summary(fm, which = "transition") 
@

@@

Multivariate data can be modelled by providing a list of formulae as
well as a list of family objects for the distributions of the various
responses.  In above examples we have only used the response times
which were modelled as a Gaussian distribution.  The accuracy
variable in the \code{speed} data can be modelled with a multinomial
by specifying the following:
<<>>=
set.seed(1)
mod <- depmix(list(rt ~ 1,corr ~ 1), data = speed, nstates = 2, 
  family = list(gaussian(), multinomial("identity")),
  transition = ~ scale(Pacc), instart = runif(2))
fm <- fit(mod,verbose = FALSE)
@

This provides the following fitted model parameters (only the 
response parameters are given here): 
<<>>=
summary(fm, which = "response")
@

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[t!]
\begin{center}
	\includegraphics{baldist.pdf}
	\caption{Balance scale item; this is a distance item (see the text 
	for details).}
	\label{fig:balance}  
\end{center}
\end{figure}
